Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis

### 2018-06-15 jhir/latex-manuscript/manuscript.tex
update manuscript based on april's feedback

Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations~\cite{bakeman1997observing, bakeman2011sequential}. Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN. A follow-up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}. Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk. Gaume et al.~\cite{gaume2008counsellor} used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service. They found \hl{that} MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed. A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk~\cite{gaume2010counselor}. In this sample, MIIN led to greater sustain talk, but was unrelated to change talk. Further analyses revealed that reflections were the only MICO behavior linked to increased change talk; reflections and other MICO behaviors, excluding questions, were related to increased sustain talk. Glynn and colleagues~\cite{glynn2014change} linked reflections of change talk to the elicitation of change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use. In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk and statements emphasizing decision-making autonomy. A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion \hl{that} asking questions phrased to elicit change talk, reflections of change talk and autonomy supportive statements \hl{are} the counselor behaviors, \hl{which lead} to the elicitation of change talk. Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in \hl{which various} MICO strategies are effective.

@@

\section{Discussion}
\label{sec:discussion}
This study applied HMM and frequent pattern mining to test the fundamental hypothesis guiding Motivational Interviewing, which posits that counselors use of ``MI-consistent'' communication strategies (MICO) will lead to patient change talk~\cite{miller2009toward}. Previous studies have empirically linked counselors' use of MICO to higher rates of patient change talk in first-order Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}. The present study leveraged data mining methods to provide an even stronger evidence for MI's fundamental hypothesis by considering longer-term dependencies in the data. Unlike simple first-order Markov chain models, frequent pattern mining considers behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement, while HMM identifies groups of communication behaviors occurring in successful and unsuccessful communication sequences. The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing, which has previously relied upon simple Markov chain models~\cite{moyers2006therapist, moyers2009session, gaume2008counsellor, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building}.

In both analyses, MICO communication strategies were characteristic of successful sequences (i.e., those resulting in a change talk statement). In HMM, the majority of successful sequences began in the \textit{``high motivation''} state, when counselors frequently use reflections of change talk or commitment language as well as affirmations. Other high frequency counselor behaviors observed in successful sequences included statements emphasizing patients' decision-making autonomy, questions phrased to elicit change talk and the provision of information using patient-centered strategies. The frequent pattern mining results were similar. Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient's decision-making autonomy and sensitively provided information. Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building} and statements emphasizing the patient's decision-making autonomy~\cite{carcone2013provider, jacques2016building}. However, these studies did not find a link between change talk and counselors' use of affirmations or the provision of information, when examined independently of the MICO index. Thus, the present study is the first to provide empirical evidence for these causal linkages. One reason for this unique finding may be the treatment context, adolescent patients engaged in a voluntary weight loss trial. Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections and the provision of information are critical communication skills for encouraging health-related behavior change~\cite{douaihy2015motivational}. Thus, providing information in a patient-centered manner in the context of health care treatment may be necessary to ensure patients have the requisite knowledge of their health care problem and its treatment.

>>>>

Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations~\cite{bakeman1997observing, bakeman2011sequential}. Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN. A follow-up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}. Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk. Gaume et al.~\cite{gaume2008counsellor} used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service. They found \hl{that} MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed. A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk~\cite{gaume2010counselor}. In this sample, MIIN led to greater sustain talk, but was unrelated to change talk. Further analyses revealed that reflections were the only MICO behavior linked to increased change talk; reflections and other MICO behaviors, excluding questions, were related to increased sustain talk. Glynn and colleagues~\cite{glynn2014change} linked reflections of change talk to the elicitation of change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use. In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk and statements emphasizing decision-making autonomy. A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion \hl{that} asking questions phrased to elicit change talk, reflections of change talk and autonomy supportive statements \hl{were} the counselor behaviors, \hl{which led} to the elicitation of change talk. Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors\hl{, however, }led to change talk in some \hl{treatment contexts} but not others suggesting a need for additional research to understand the \hl{treatment contexts in which various} MICO strategies are effective. \hl{The current study contributes to existing knowledge by examining African American adolescents in weight loss treatment.}

@@

\section{Discussion}
\label{sec:discussion}
This study applied HMM and frequent pattern mining to test the fundamental hypothesis guiding Motivational Interviewing, which posits that counselors use of ``MI-consistent'' communication strategies (MICO) will lead to patient change talk~\cite{miller2009toward}. Previous studies have empirically linked counselors' use of \hl{MICO communication strategies} to higher rates of patient change talk in first-order Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}. The present study leveraged data mining methods to provide an even stronger evidence for MI's fundamental hypothesis by considering longer-term dependencies in the data. Unlike simple first-order Markov chain models, frequent pattern mining considers behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement, while HMM identifies groups of communication behaviors occurring in successful and unsuccessful communication sequences. The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing, which has previously relied upon simple Markov chain models~\cite{moyers2006therapist, moyers2009session, gaume2008counsellor, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building}.

In both analyses, MICO communication strategies were characteristic of successful sequences (i.e., those resulting in a change talk statement). In HMM, the majority of successful sequences began in the \textit{``high motivation''} state, when counselors frequently use reflections of change talk or commitment language as well as affirmations. Other high frequency counselor behaviors observed in successful sequences included statements emphasizing patients' decision-making autonomy, questions phrased to elicit change talk and the provision of information using patient-centered strategies. The frequent pattern mining results were similar. Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient's decision-making autonomy and sensitively provided information. Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building} and statements emphasizing the patient's decision-making autonomy~\cite{carcone2013provider, jacques2016building}. However, these studies did not find a link between change talk and counselors' use of affirmations or the provision of information, \hl{when examining specifically which of the MICO communication strategies were empirically linked to the elicitation of change talk.} Thus, the present study is the first to provide empirical evidence for these causal linkages. One reason for this unique finding may be the treatment context, adolescent patients engaged in a voluntary weight loss trial. Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections and the provision of information are critical communication skills for encouraging health-related behavior change~\cite{douaihy2015motivational}. Thus, providing information in a patient-centered manner in the context of health care treatment may be necessary to ensure patients have the requisite knowledge of their health care problem and its treatment.
